---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The variables included in this dataset are:

1. steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. date: The date on which the measurement was taken in YYYY-MM-DD format and stored as date in the dataframe
3. interval: Identifier for the 5-minute interval in which measurement was taken and stored as numeric in the dataframe

```{r, echo=TRUE}
library(lattice)

DataSet <- read.csv("activity.csv", sep=",", header=TRUE, stringsAsFactors=FALSE)
DataSet[,"date"] <- as.Date(DataSet[,"date"], "%Y-%m-%d")
DataSet$interval <- as.numeric(DataSet$interval)

```


## What is mean total number of steps taken per day?

For this part of the assignment, the missing values in the dataset are ignored. However, values recorded as 0 are included for mean and median computation.

#### Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}
TotalByDay <- aggregate(DataSet[, "steps"], list(DataSet[,"date"]), sum, na.rm=TRUE) 
colnames(TotalByDay) <- c("date", "stepsTotal")
hist(TotalByDay$stepsTotal, breaks=40, ylim=c(0,12), xlim=c(0,25000), xlab="Total Steps per day", ylab="No of days", main="Distribution of total steps/day during Oct & Nov 2012", col="blue")
```

#### Calculate and report the mean and median total number of steps taken per day

```{r, echo=TRUE}
stepsTotalMean <- round(mean(TotalByDay$stepsTotal, na.rm=TRUE),2)  
stepsTotalMedian <- median(TotalByDay$stepsTotal, na.rm=TRUE) 
```
- `r  stepsTotalMean` is the Mean of total number of steps taken per day

- `r stepsTotalMedian` is the Median of total number of steps taken per day

## What is the average daily activity pattern?

#### Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
MeanByInterval <- aggregate(DataSet[, "steps"], list(DataSet[,"interval"]), mean, na.rm=TRUE) 
colnames(MeanByInterval) <- c("interval", "stepsinterval")
plot(MeanByInterval$interval, MeanByInterval$stepsinterval, type="l", ylim=c(0,210), xlim=c(0,2355), ylab="Steps per interval", xlab="Interval", main="Average steps/interval during Oct & Nov 2012", col="blue", axes=FALSE)
axis(side=1, at=seq(0,2400, 200), labels=seq(0,2400,200))
axis(side=2, at=seq(0,225, 25), labels=seq(0,225,25))
MeanByInterval <- MeanByInterval[order(MeanByInterval$stepsinterval, decreasing=TRUE),]
````

- `r MeanByInterval[1,"interval"] ` is the 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps
- `r round(MeanByInterval[1,"stepsinterval"],2) ` is the corresponding number of steps for the above interval

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
- `r nrow(DataSet[is.na(DataSet$steps),]) ` 

#### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc. Create a new dataset that is equal to the original dataset but with the missing data filled in.
- The strategy used is copying 5-minute interval average to cells with NA.

```{r, echo=TRUE}
DataSetNew <- merge( MeanByInterval, DataSet, by="interval")
for (i in 1:nrow(DataSetNew)) {
	DataSetNew[i,"steps"] <- if (is.na(DataSetNew[i,"steps"])) { DataSetNew[i,"stepsinterval"]} else {DataSetNew[i,"steps"]}
	}
```
#### Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}
TotalByDayNew <- aggregate(DataSetNew[, "steps"], list(DataSet[,"date"]), sum, na.rm=TRUE) 
colnames(TotalByDayNew) <- c("date", "stepsTotal")
hist(TotalByDayNew$stepsTotal, breaks=40, ylim=c(0,20), xlim=c(0,50000), xlab="Total Steps per day", ylab="No of days", main="Distribution of total steps/day during Oct & Nov 2012 - after Imputing", col="green")
```

#### Calculate and report the mean and median total number of steps taken per day. 

```{r, echo=TRUE}
stepsTotalMeanNew <- round(mean(TotalByDayNew$stepsTotal, na.rm=TRUE),2)  
stepsTotalMedianNew <- median(TotalByDayNew$stepsTotal, na.rm=TRUE) 
```
- `r as.character(round(stepsTotalMeanNew,2), "99999.99")` is the Mean of total number of steps taken per day

- `r as.character(round(stepsTotalMedianNew,2), "99999.99")` is the Median of total number of steps taken per day

##### Do these values differ from the estimates from the first part of the assignment? 
- The mean has increased by `r round( (stepsTotalMeanNew - stepsTotalMean)* 100/stepsTotalMean, 2)` percent
- The median has decreased by `r abs(round((stepsTotalMedianNew - stepsTotalMedian)* 100/stepsTotalMedian, 2))` percent

##### What is the impact of imputing missing data on the estimates of the total daily number of steps?
- The summary statistics shown below indicate a more balanced distribution after imputing

```{r, echo=TRUE}
summary(TotalByDay$stepsTotal)
summary(TotalByDayNew$stepsTotal)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
DataSet$WeekDay <- sapply(weekdays(DataSet$date,abbreviate = TRUE), switch, 
 "Mon" = "weekday",
 "Tue" =  "weekday",
 "Wed" =  "weekday",
 "Thu" =  "weekday",
 "Fri" =  "weekday",
 "Sat" =  "weekend",
 "Sun" =  "weekend"
)

DataSet$WeekDay <- as.factor(DataSet$WeekDay)

MeanByIntervalWeekday <- aggregate(DataSet[, "steps"], list(DataSet[,"interval"], DataSet[,"WeekDay"]), mean, na.rm=TRUE) 
colnames(MeanByIntervalWeekday) <- c("interval", "WeekDay", "stepsMean" )
xyplot(stepsMean ~ interval | WeekDay, data = MeanByIntervalWeekday, layout = c(1, 2), type="l", ylab="Number of steps")

````
